Double White Dwarf Mergers with CASTRO\\ I. Methodology and Code Verification

### 2015-01-16 papers/wdmerger_I/paper.tex
some appendix updates

The basic PPM algorithm in CASTRO has undergone a number of changes since
the original code paper \citep{castro}.  A discussion of the pure
hydrodynamics changes was given in \citet{zingalekatz:2015}.  Here
we discuss the changes that affect multispecies flow and source
terms.

@@

We use the Castro code as described in \citet{castro}.  For all the
runs, the PPM reconstruction is done, using the original limiters for
the parabolic profiles \citep{ppm}.  We modify the prediction of the
interface states slightly.  The original implementation of the PPM
prediction in Castro takes the form:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = q_i -
   \sum_{\nu;\lambda_i^{(\nu)}\ge 0} l_i^{(\nu)} \cdot \left [
        q_i - \mathcal{I}_+^{(\nu)}(q_i)
       \right ] r_i^{(\nu)}
\end{equation}
where $q$ is the vector of primitive variables, with $q_i$
representing the average in the cell, $l^{(\nu)}$ and $r^{(\nu)}$ are
the left and right eigenvectors with eigenvalue $\lambda^{(\nu)}$,
with $\nu$ the index of the characteristic wave of the system.  The
sum is over all the waves that result from the characteristic
structure of the problem, but designed such that only waves moving
toward the interface contribute to the interface value,
$q_{i+1/2,L}^{n+1/2}$.  Finally, $\mathcal{I}_+^{(\nu)}(q)$ is the
average under the parabolic profile of quantity $q$ of all the information
that can reach the right interface of the zone $i$ as carried by the wave
$\nu$.   The reader is referred to
\citet{ppmunsplit} for further defaults.

@@

The presence of $q_i$ in this expression serves as a reference state.
The idea is that the error in the characteristic projection of the
jump (due to the nonlinearity) is minimized if we pick a suitable
reference state, since we only care about the jumps carried to the
interface over the timestep \citep{colellaglaz1985}.  The original
Castro implementation used the cell-average quantity.  Experiments
found that this is ill-behaved in the presence of strong shocks
(especially at low resolution). For the present work, we switch the
reference state to
\begin{equation}
\label{eq:refchoice}
\tilde{q}_L = \left \{ \begin{array}{cc}
       \mathcal{I}_+^{(+)}(q_i) & \mathrm{if~} u + c > 0 \\
       q_i                    & \mathrm{otherwise}
\end{array}
\right .
\end{equation}
where the $(+)$ superscript here means the fastest wave moving to the right
(the $u+c$ eigenvalue).   This is simply the average under the largest
portion of the parabolic profile that could possible reach the interface 
over the timestep.  This is
in agreement with \citet{ppmunsplit} (eq. 90).  This makes our
expression appear as:
\begin{equation}
\label{eq:ppmstatel}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L -
   \sum_{\nu;\lambda_i^{(\nu)}\ge 0} l_i^{(\nu)} \cdot \left [
        \tilde{q}_L  - \mathcal{I}^{(\nu)}_+(q_i)
       \right ] r_i^{(\nu)}
\end{equation}
The final change is that now we can no longer simply multiply
$\tilde{q_i} - \mathcal{I}^{(\nu)}_+(q_i)$ by the flattening
parameter, $\chi$, but instead blend the traced state with the cell
centered state as:
  \begin{equation}
  q_{i+1/2,\{L,R\}}^{n+1/2} \leftarrow (1 - \chi) q_i + \chi q_{i+1/2,\{L,R\}}^{n+1/2}
  \end{equation}
This is the same prescription used in the \cite{ppm}.

This change is now the default in the public version of Castro,
and can be controlled with the parameter {\tt castro.ppm\_reference}.
Figure~\ref{Fig:sod} shows the
solution to the Sod problem using the old and new reference states.
We see that the shock and contact are slightly sharper with this new
reference state, but also there is a slight dip at the tail of the
rarefaction with the new method.

It is instructive to look at the effect of the reference state choice.
First, consider one of the variables present in one-dimensional flow
(density, velocity in the normal direction, and pressure).  Here we
our reference state is as in Eq.~\ref{eq:refchoice} and we have.
Ignoring flattening, if there are no waves moving toward our
interface, then Eq.~\ref{eq:ppmstatel} reduces to:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L = q_i
\end{equation}
If instead only the fastest wave is moving toward the interface, then
only the term corresponding to the fastest wave in the sum will be
added in Eq.~\ref{eq:ppmstatel}, but our choice of reference state makes
that term 0 by design, and our interface state is:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L = \mathcal{I}_+^{+}(q_i)
\end{equation}
This is the desired behavior for each of these cases.

>>>>

The basic PPM algorithm in CASTRO has undergone a number of changes
since the original code paper \citep{castro}.  A discussion of the
pure hydrodynamics changes along with verification of CASTRO when
using the stellar equation of state was given in
\citet{zingalekatz:2015}.  Here we discuss the changes that affect
multispecies flow and source terms.

@@

@@

We use the Castro code as described in \citet{castro}.  For all the
runs, the PPM reconstruction is done, using the original limiters for
the parabolic profiles \citep{ppm}.  The prediction of the interface
states appears as:
\begin{equation}
\label{eq:ppmstatel}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L -
   \sum_{\nu;\lambda_i^{(\nu)}\ge 0} l_i^{(\nu)} \cdot \left [
        \tilde{q}_L  - \mathcal{I}^{(\nu)}_+(q_i)
       \right ] r_i^{(\nu)}
\end{equation}
where $q$ is the vector of primitive variables, $l_i^{(\nu)}$ and
$r_i^{(\nu)}$ are the left and right eigenvectors with eigenvalue
$\lambda_i^{(\nu)}$, with $\nu$ the index of the characteristic wave of
the system.  The sum is over all the waves that result from the
characteristic structure of the problem, but designed such that only
waves moving toward the interface contribute to the interface value,
$q_{i+1/2,L}^{n+1/2}$.  The reference state, $\tilde{q}_L$ is
chosen to minimize the work of the characteristic projection.
Finally, $\mathcal{I}_+^{(\nu)}(q)$ is the
average under the parabolic profile of quantity $q$ of all the
information that can reach the right interface of the zone $i$ as
carried by the wave $\nu$.  The reader is referred to
\citet{ppmunsplit} for further defaults.

Since the original CASTRO paper, the reference state implementation
has been switched to:
\begin{equation}
\label{eq:refchoice}
\tilde{q}_L = \left \{ \begin{array}{cc}
       \mathcal{I}_+^{(+)}(q_i) & \mathrm{if~} u + c > 0 \\
       q_i                    & \mathrm{otherwise}
\end{array}
\right .
\end{equation}
where the $(+)$ superscript here means the fastest wave moving to the right
(the $u+c$ eigenvalue).   This is simply the average under the largest
portion of the parabolic profile that could possible reach the interface 
over the timestep.  This is
in agreement with \citet{ppmunsplit} (eq. 90).  The flattening
in the original CASTRO paper has also been updated as discussed
in \citet{zingalekatz:2015}.

We comment on the choice of reference state for passively-advected
quantities (like $X_k$ or the transverse velocity).
First, consider one of the variables present in one-dimensional flow
(density, velocity in the normal direction, and pressure).  Here we
our reference state is as in Eq.~\ref{eq:refchoice} and we have.
Ignoring flattening, if there are no waves moving toward our
interface, then Eq.~\ref{eq:ppmstatel} reduces to:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L = q_i
\end{equation}
If instead only the fastest wave is moving toward the interface, then
only the term corresponding to the fastest wave in the sum will be
added in Eq.~\ref{eq:ppmstatel}, but our choice of reference state makes
that term 0 by design, and our interface state is:
\begin{equation}
q_{i+1/2,L}^{n+1/2} = \tilde{q}_L = \mathcal{I}_+^{+}(q_i)
\end{equation}
This is the desired behavior for each of these cases.
